\hspace{-0.1cm}{J-ReCoVer}:~Java~Reducer~Commutativity~Verifier

### 2018-11-14 draft.tex
polish

\title{J-ReCoVer: Java Reducer Commutativity Verifier}

@@

In this paper, we propose heuristics for the commutativity problem that works very efficiently on a large set of practical integer reducer programs. We reduce the commutativity problem to a SMT problem in a sounds but incomplete manner; it  reports commutative only when it holds for all possible initial valuations, even if some of them are in fact unreachable. We also propose optimizations to enable the approach working on real world examples. For the case that the reducer is not proven commutative, we complement the approach by using testing to find concrete counterexamples. We implement those heuristics as a tool called J-ReCoVer (Java Reducer Commutativity Verifier), which is available at~\verb|http://jrecover.iis.sinica.edu.tw|.

@@

\subsection*{Related Works}
Reducer commutativity problem can be reduced to program equivalence problem. One creates another program $R'$ that first non-deterministically swaps two consecutive input values and then executes the code of $R$. If $R'$ and $R$ are equivalent, using the fact that all permutations of a list can be obtained by swapping consecutive list elements finitely many times\footnote{Here is an example to produce [3;2;5;1;4] from [1;2;3;4;5] by swapping consecutive elements: [\textbf{1};\textbf{2};3;4;5] $\rightarrow$ [2;\textbf{1};\textbf{3};4;5] $\rightarrow$ [2;3;1;\textbf{4};\textbf{5}]$\rightarrow$ [2;3;\textbf{1};\textbf{5};4]$\rightarrow$ [\textbf{2};\textbf{3};5;1;4]$\rightarrow$ [3;2;5;1;4].}, $R$ can be proved to be commutative. There are a series of research results on program equivalence checking (or closely related topics such as regression verification)~\cite{symdiff,fedyukovich2015automated,sharma2013data,godlin2009regression,fedyukovich2016property,felsing2014automating,KieferKlebanovUlbrich2017,lahiri2013differential,grossman2017verifying,barthe2011relational,KlebanovRuemmerUlbrich2017}. At a high level, checking equivalence of two programs $P$ and $P'$ can be reduced to a sequential verification problem by executing $P'$ after $P$ and at the end checking if they always have the same outputs. It can be made more efficient by finding the right~\emph{synchronization points} and combining the code of $P$ and $P'$ in an interleaved manner. A lot of the research effort in this research direction is in finding good synchronization points. For the case of commutativity analysis, good synchronization points are already given; one can simply use the beginning of the loop as the synchronization point. However, in our experience, if we naively reduce the commutativity problem to equivalence problem and check it in a precise manner, many examples cannot be solved. In this paper, we propose a good over-approximation (c.f. Section~\ref{sec:prover}) of the reducer's behavior. The approximation allows a much more efficient yet precise enough commutativity analysis.

One main motivation of commutativity analysis is to reduce verification problem of a concurrent program to a sequential one. There are several researches sharing the similar motivation. For instance, the concept of \emph{robustness} of event-driven asynchronous programs~\cite{ahmed2017:robustness}  and program running under relaxed memory model~\cite{ahmed2013:robustness,AbdullaACLR13,AbdullaACLR12}, \emph{synchronizability} of communicating finite state machines~\cite{FinkelL17,basu2012synchronizability,basu2011choreography}. However, their computation models are quite different to that of reducers and hence the result cannot be directly applied. Besides verification, another interesting research direction is the synthesis of MapReduce programs~\cite{SmithA16}. Commutativity analysis is used as a component under the synthesis framework.

\section{Notations and Definitions}
\label{section:integer-reducers}
We use $\Z$ to denote the set of integers and $[n,m]$ to denote the set of integers $\{k\mid n \leq k\leq m\}$. 
We lift the equality $=$ operation to tuples in the standard way.
Let $S\subseteq [1,k]$, the equality $\{v_1,\ldots,v_k\}=_{S} \{v'_1,\ldots,v'_k\}$ between tuples holds if and only if $\bigwedge_{i\in S} v_i=v'_i$.
For presentation purpose, we describe a highly simplified model to characterize the essence of reducers. The model allows to describe the computation of reducers and investigate their commutativity.

@@

Here $v\in \Var$ and $c \in \Z$ are variables and constant integer values, respectively. An \emph{integer expression} in $\Exp$ can be either a variable, constant value, a call to the $\cur$ function that reads and consumes an input value of the reducer, their combinations over basic arithmetic operations, or a non-deterministically chosen integer value $*$.
A \emph{guard} in $\Grd$ can be a comparison between two expressions or boolean combinations of several guards. A \emph{command} in $\Cmd$ can be an assignment, a branch statement, concatenation of a sequence of commands, or $out(v)$ that outputs the value of $v$.
A \emph{reducer program} is defined as $s_1;\rloop\{s_2\};s_3$, where $s_1,s_2,s_3 \in \Cmd$, and $s_2$ is assumed to have at most one $\cur$ expression in each branch\footnote{We define a \emph{branch} of a command $s$ as a sequence obtained by replacing every branch statement $\ite{g}{s_1}{s_2}$ in $s$ with either $(g)\wedge s_1$ or $(not\ g) \wedge s_2$.} . According to our observation over hundreds  of reducer programs in open repositories, reducer programs are almost always in this form. The $\rloop\{s_2\}$ statement enters the loop body to execute $s_2$ repeatedly until the entire input list is consumed. A few examples of reducer programs in this model can be found in Figure~\ref{fig:reducer_example}.

Some reducer uses two loops to compute the output. For example, for calculating standard deviation, one first computes the average of inputs and then uses it to compute the standard deviation. In that case, we can still verify the reducer by first partitioning it into two reducers and then verifying them separately. One reducer computes the average (Figure~\ref{fig:reducer_example}(b)) and the other reducer uses a non-deterministic value to over-approximate the computed average ($a := *$) and then use it in the computation of standard deviation (Figure~\ref{fig:reducer_example}(c)). The reducer in Figure~\ref{fig:reducer_example}(c) ``over-approximates''  all possible output of the standard deviation computation; it computes an output even if $a$ is not the average of inputs. However, the reducer is still commutative and its commutativity verification can still be done.

@@

\textbf{The Preprocessor component} first compiles a reducer program to bytecode and use the tool Soot~\cite{soot} to further convert it to the so called ``Jimple'' format, which is an intermediate language that is designed to simplift the analysis of Java programs. Under the Hadoop MapReduce framework, the permutation of the input is handled by the scheduler/shuffler component and is affected by things like network latency, which is not controllable by programmers. In order to deal with this issue, we wrote our own dummy Hadoop environment for the reducer as a part of the Preprocessor component. So that the input order of the reducer is now controlled by J-ReCoVer. Finally, we perform in the Preprocessor some simple program transformations tasks to simplify the analysis. For example, we convert all output commands $out(e)$ inside loops to an special assignment (Section~\ref{sec:program_trans1} and~\ref{sec:program_trans2}).

@@

\textbf{The Prover component} reduces the commutativity problem to a SMT formula satisfiability problem. At a high level, we check equivalence between a reducer program and its variant that have two consecutive inputs swapped. We show that this can be reduced to a SMT problem and give it to SMT solver Z3~\cite{z3} for solving. Recall that all permutations of a list can be obtained by swapping consecutive elements finitely many times. In case that Z3 proved the formula unsatisfiable, we know that swapping any two consecutive inputs of the reducer will not change its output. It follows that the reducer outputs the same value for all permutations of the same list of input.
In this case,  J-ReCoVer stops and reports that the reducer is commutative.

If we run the approach in a naive manner, it does not scale to large programs and is very imprecise.  We will introduce two critical optimizations that enables the proposed approach working for real-world examples in Section~\ref{section:optimizations}: a data flow analysis to extract critical variables and an improved symbolic execution algorithm to generate Z3 formulae. More details of the algorithms used in the Preprocessor and the Prover components will be explained in Section~\ref{sec:preprocessor_prover} and the optimizations will be introduced in~\ref{section:optimizations}.

@@

Before entering the Prover component, the Preprocessor first performs a few program transformation tasks to simplify the verification task. 
We assume the reducer $s_1;\rloop\{s_2\};s_3$ is given. We begin the section with our first program transformation task (Section~\ref{sec:program_trans1}) that handles the case when $s_2$ contains $out$ commands. The output of the first transformation task is a commutativity-equivalent reducer program where $out$ does not occur in $s_2$.  Then we explain the algorithm of the Prover (Section~\ref{sec:prover}) that works only in a special case where no read from input occurs in $s_1$. For general reducer programs, we introduce the second program transformation task (Section~\ref{sec:program_trans2}) whose output is a reducer program that contains all input/output behaviors of the original reducer (may contain behaviors that are not in the original reducer) and there is no read from input (i,e., does not have the $\cur$ expression) in $s_1$, which can be analyzed by the algorithm of Prover.

@@

\subsection{The Prover: Reducing Commutativity to SMT Solving}
\label{sec:prover}
Below we begin with a simple version where $s_1$ never calls the $\cur$ function. Later we will explain how to handle the general case using program transformation. The command $s_2$ can be viewed as a function $F$ that reads the values of all variables and the current input before executing $s_2$ and outputs values of all variables after $s_2$. Formally, the function $F(n,x_1,x_2,\ldots,x_k): \Z^{k+1} \rightarrow \Z^k$ returns a tuple of values $x_1',x_2',\ldots,x_k'$, where $k$ is the current input value of the reducer, $x_i$ and $x_i'$ are the values of variables before and after the execution of $s_2$, respectively. The construction of $F$ from $s_2$ can be done in the standard way. We will introduce a more efficient data structure for the construction in Section~\ref{section:optimizations}.

We check if the formula below is valid for all possible values of $n_1,n_2, x_1,x_2,\ldots,x_k$. 
\begin{equation}
 F(n_1, F(n_2,x_1,x_2,\ldots,x_k)) = F(n_2, F(n_1,x_1,x_2,\ldots,x_k) )
\label{eq:commu}
\end{equation}

@@

The validity of Equation~(\ref{eq:commu}) implies that swapping any two consecutive input values does not change the final variable valuation. Informally, it says that starting from the same initial valuations of variables and with two different input orders, the values of all program variables are the same after we execute the loop body twice. The first execution reads $n_1$ and then reads $n_2$. The other execution reads the two inputs in the reverse order.

@@

We explain why that the validity of Equation~(\ref{eq:commu}) is sufficient to prove the commutativity of the reducer $s_1;\rloop\{s_2\};s_3$. Notice that $s_1$ does not read from the input (assumed at the beginning of this subsection). The execution of $s_1$ is deterministic in the sense that from the same initial valuation, it outputs the same sequence of values and ends at the same valuation. For the execution of $\rloop\{s_2\}$, from the validity of Equation~(\ref{eq:commu}), any permutation of input values leads the reducer to the same final valuation and the same sequence of outputs at the end of the loop. 
Before executing the commands after the loop $s_3$, all inputs have been read (the loop exit condition) already, similar the case of $s_1$, the execution of $s_3$ is deterministic for all permutations of inputs and outputs the same sequence of values.

@@

We handle the problem again by program transformation. To handle the example in Figure~\ref{fig:reducer_max} (a), we move the prefix $s_1$ into the loop body and use a new variable $s$ to force that $s_1$ is always executed before the original loop body. The result after the transformation is demonstrated in Figure~\ref{fig:reducer_max} (b). Observe that the new reducer program over-approximates all possible inputs/outputs of the original one.
It also satisfies all assumptions required for the Prover component: (a) The commands before entering the loop does not have any $\cur$ expression and (II) each branch (recall that ``branch'' is formally defined in Section~\ref{section:integer-reducers}) of the loop body has at most one $\cur$. So we can apply Equation~(\ref{eq:commu}) to prove its commutativity.
Since the new reducer has all the possible outputs of the original one, the new reducer being commutative implies that the original one is also commutative.

\begin{figure}[htb]
	\centering
	\begin{minipage}{0.8\textwidth}
	\begin{algorithm}[H]
		$s'_0;s:=*;$\;
		\Loop{}{
			\lIf{$s=1$}{ $x_{i_1} := \cur; s'_1; s:=2$}
			\lElseIf{ $s =2$}{ $x_{i_2} := \cur; s'_2; s:=3$}
			\lElseIf{ $s = 3$}{$\ldots$}
			\lElseIf{ $s = m$}{$x_{i_m} := \cur; s'_m; s:=m+1$}
			\lElse{ $s_2$}
		};$s_3$\;
	\end{algorithm}
	\end{minipage}
	\caption{The second transformation task}
	\label{fig:general_program_transformation}
\end{figure}

For the general case, we first rewrite the $s_1$ part of the reducer $s_1;\rloop\{s_2\};s_3$ to the form 
$$s'_0;x_{i_1}{:=} \cur;s'_1.\ldots x_{i_{m}}{:=}\cur;s'_m, $$ where $\cur$ never occur in $s'_0,s'_1,\ldots, s'_m$. Then we produce the reducer in Figure~\ref{fig:general_program_transformation}, which is an over-approximation of the original reducer. For the case that $s=1$ at the beginning, its output sequence is the same as the original reducer. When $s\neq 1$, it may produce outputs that are not possible by the original reducer. Observe that the reducer in Figure~\ref{fig:general_program_transformation} satisfies all assumptions required for the Prover: each branch has at most one $\cur$. Hence the Equation~(\ref{eq:commu}) can then be applied to prove its commutativity.

\section{Optimizations}
\label{section:optimizations}
In this section we explain how data flow analysis are used in J-Recover to improve the precision of commutativity checking and the algorithm we used to construct the $F$ function in Equation~(\ref{eq:commu}).

\subsection{Improving the precision of commutativity verification using data flow analysis}
\label{section:opt1}
We find that in many cases, Equation~(\ref{eq:commu}) is a condition that is too strong for the reducer to be commutative. For instance, consider the example in Figure~\ref{fig:reducer_opt}, in the loop body, the input is first stored in a variable $t$ and then added to $s$. In this case, the $F$ function returns the updated values of both $s$ and $t$ after the execution of loop body. Observe that 
$F(c_1, F(c_2,s,t) = (c_1+c_2, c_1)$ and $F(c_2, F(c_1,s,t) = (c_1+c_2, c_2)$. It follows that the corresponding Equation~(\ref{eq:commu}) is invalid. The second component of the returned tuple, which corresponds to the value of $t$ after executing the loop body twice, are $c_1$ and $c_2$, respectively. However, in this case, the value of $t$ is not important to the output of this reducer.

@@

To handle such a situation, we perform a simple backward data flow analysis to collect all variables whose value may propagate to the output command after the loop execution. With out loss of generality, we assume there is only one output command after the loop. Then the code after the loop (i.e., the $s_3$ part of the reducer) can only be in the form of Figure~\ref{fig:nested_out}.
We unfold all concatenation of commands, so now the commands $c_1\ldots c_n$ are either assignment commands $w:=e$ or branch command $\ite{g}{c'_1}{c'_2}$, and the only output command stays in the $m$-th layer of the branch command after $c_i$. For the case that $m=0$, the entire branch command in Lines 2-6 of Figure~\ref{fig:nested_out} is replaced with $out(v)$.

The data flow analysis algorithm works as follows. Recall that $s_3$ can only be in from of Figure~\ref{fig:nested_out}.
A set $V_3$ whose value is initially a singleton set $\{v\}$ is used to remember important variables that might flow to the output command $out(v)$ in $s_3$.
The commands and guards will be inspected in the following order $c_{n}$,$\ldots$,$c_{i+1}$, $g_m$, $g_{m-1}$, $\ldots$, $g_1$, $c_i$, $\ldots$, $c_2$, $c_1$. For the case that a guard $g_j$ for $j\in [1,m]$ is being processed, we simply add all variables occurred in $g_j$ to $V_3$. 
For the case that a command $c_j$ for $j\in [1,n]$ is being processed, 
recall that $c_j$ can only be in the form of an assignment $w:=e$ or a branch $\ite{g}{c'_1}{c'_2}$.
For the case that $c_j$ is an assignment $w:=e$ and $w\in V_3$, we add all variables contained in $e$ to $V_3$. For the case that $c_j$ is a branch command $\ite{g}{c'_1}{c'_2}$, we process both $c'_1$ and $c'_2$ in a similar manner. Starting from the same set $V_3$,  after processing $c'_1$ and $c'_2$, we get respectively two sets $V_3^1$ and $V_3^2$, we take their union $V_3^1\cup V_3^2$ as the new value of $V_3$. After the entire $s_3$ is being processed we get the set of important variables that might affect the outputs.
We define the set $V_2$ as the set of variables that the first program transformation task used to collect output values in the loop body $s_2$, i.e., $V_2= \{v_{out_k} \mid k \in [1,l]\}$, where $l$ is the number of output commands occurred in $s_2$.

Then we modify the Equation~(\ref{eq:commu}) to the one below
\begin{equation}
F(n_1, F(n_2,x_1,x_2,\ldots,x_k) )=_{PosF(V_2\cup V_3)} F(n_2, F(n_1,x_1,x_2,\ldots,x_k) )
\label{eq:commu2}
\end{equation}
The function $PosF(X)$ returns the corresponding positions of variables in $X$ in the output of $F$. E.g., if $X = \{x_1,x_4,y\}$, where $y\notin \{x_1,x_2,\ldots,x_n\}$, then $PosF(X)=\{1,4\}$.
The use of equation~(\ref{eq:commu2}) is critical to the success of our approach. In our evaluation, the ratio of examples that our approach can successfully analyze is significantly increased from $6.8\%$ to $97.5\%$. The details of the evaluation can be found in Section~\ref{section:exp}.

@@

\subsection{Comparison with competing tools}
As already pointed out in~\cite{ChenHSW15}, a reducer commutativity problem can be reduced to a reachability problem of a program. 
The examples in~\cite{ChenHSW15}, which includes \textsf{rangesum}\footnote{The sum of the second half of the input list.}, \textsf{avg}, \textsf{max}, \textsf{sep}\footnote{The number of occurrences of even numbers minus that of odd numbers.}, \textsf{sum}, are already included in SV-Comp~\cite{svcomp} (Software Verification Competition) benchmarks.
Although those examples are tiny and simple, none of the software verifiers are able to verify all of them according to the 2018 results. In contrast, J-Recover can verify all them very efficiently.

>>>>

\title{\hspace{-0.1cm}{J-ReCoVer}:~Java~Reducer~Commutativity~Verifier}

@@

In this paper, we propose heuristics for the commutativity problem that works very efficiently on a large set of practical integer reducer programs. We reduce the commutativity problem to a SMT problem in a sound but incomplete manner. We also propose optimizations to enable the approach working on real world examples. For the case that the reducer is not proven commutative, we complement the approach by using testing to find concrete counterexamples. We implement those heuristics as a tool called J-ReCoVer (Java Reducer Commutativity Verifier), which is available at~\verb|http://jrecover.iis.sinica.edu.tw|.

@@

\subsection*{Related Works}
Reducer commutativity problem can be reduced to program equivalence problem. One creates another program $R'$ that first non-deterministically swaps two consecutive input values and then executes the code of $R$. If $R'$ and $R$ are equivalent, using the fact that all permutations of a list can be obtained by swapping consecutive list elements finitely many times\footnote{Here is an example to produce [3;2;5;1;4] from [1;2;3;4;5] by swapping consecutive elements: [\textbf{1};\textbf{2};3;4;5] $\rightarrow$ [2;\textbf{1};\textbf{3};4;5] $\rightarrow$ [2;3;1;\textbf{4};\textbf{5}]$\rightarrow$ [2;3;\textbf{1};\textbf{5};4]$\rightarrow$ [\textbf{2};\textbf{3};5;1;4]$\rightarrow$ [3;2;5;1;4].}, $R$ can be proved to be commutative. There is a series of research results on program equivalence checking (or closely related topics such as regression verification and translation validation)~\cite{Pnueli:1998:TV,symdiff,fedyukovich2015automated,sharma2013data,godlin2009regression,fedyukovich2016property,felsing2014automating,KieferKlebanovUlbrich2017,lahiri2013differential,grossman2017verifying,barthe2011relational,KlebanovRuemmerUlbrich2017}. At a high level, checking equivalence of two programs $P$ and $P'$ can be reduced to a sequential verification problem by executing $P'$ after $P$ and at the end checking if they always have the same outputs. It can be made more efficient by finding the right~\emph{synchronization points} and combining the code of $P$ and $P'$ in an interleaved manner. A lot of research effort in this direction is in finding good synchronization points. For the case of commutativity analysis, one can simply use the beginning of the loop as the synchronization point. However, in our experience, if we naively reduce the commutativity problem to equivalence problem and check it in a precise manner, many examples cannot be solved. In this paper, we propose a good over-approximation (c.f. Section~\ref{sec:prover}) of the reducer's behavior. The approximation allows a much more efficient yet precise enough commutativity analysis.

One main motivation of commutativity analysis is to reduce the verification problem of a concurrent program to a sequential one. There are several researches sharing the similar motivation. For instance, the concept of \emph{robustness} of event-driven asynchronous programs~\cite{ahmed2017:robustness}  and program running under relaxed memory model~\cite{ahmed2013:robustness,AbdullaACLR13,AbdullaACLR12}, \emph{synchronizability} of communicating finite state machines~\cite{FinkelL17,basu2012synchronizability,basu2011choreography}. However, their computation models are quite different to that of reducers and hence the result cannot be directly applied. Besides verification, another interesting research direction is the synthesis of MapReduce programs~\cite{SmithA16}. Commutativity analysis is used as a component under the synthesis framework.

\section{Notations and Definitions}
\label{section:integer-reducers}
We use $\Z$ to denote the set of all integers and $[n,m]$ to denote the set of integers $\{k\mid n \leq k\leq m\}$. 
We lift the equality $=$ operation to tuples in the standard way.
Let $S\subseteq [1,k]$, the equality $\{v_1,\ldots,v_k\}=_{S} \{v'_1,\ldots,v'_k\}$ between tuples holds if and only if $\bigwedge_{i\in S} v_i=v'_i$.
For presentation purpose, we describe a highly simplified model to characterize the essence of reducers. The model allows to describe the computation of reducers and investigate their commutativity.

@@

Here $v\in \Var$ and $c \in \Z$ are variables and constant integer values, respectively. An \emph{integer expression} in $\Exp$ can be either a variable, constant value, a call to the $\cur$ function that reads and consumes an input value of the reducer, their combinations over basic arithmetic operations, or a non-deterministically chosen integer value $*$.
A \emph{guard} in $\Grd$ can be a comparison between two expressions or boolean combinations of several guards. A \emph{command} in $\Cmd$ can be an assignment, a branch statement, concatenation of a sequence of commands, or $out(v)$ that outputs the value of $v$.
A \emph{reducer program} is defined as $s_1;\rloop\{s_2\};s_3$, where $s_1,s_2,s_3 \in \Cmd$, and $s_2$ is assumed to have at most one $\cur$ expression in each branch\footnote{We define a \emph{branch} of a command $s$ as a sequence obtained by replacing every branch statement $\ite{g}{s_1}{s_2}$ in $s$ with either $(g)\wedge s_1$ or $(not\ g) \wedge s_2$.}. According to our observation over hundreds  of reducer programs in open repositories, reducer programs are almost always in this form. The $\rloop\{s_2\}$ statement enters the loop body to execute $s_2$ repeatedly until the entire input list is consumed. A few examples of reducer programs in this model can be found in Figure~\ref{fig:reducer_example}.

Some reducer uses two loops to compute the output. For example, for calculating standard deviation, one first computes the average of inputs and then uses it to compute the standard deviation. In that case, we can still verify the reducer by first partitioning it into two reducers and then verifying them separately. One reducer computes the average (Figure~\ref{fig:reducer_example}(b)) and the other reducer uses a non-deterministic value to over-approximate the computed average ($a := *$) and then use it in the computation of standard deviation (Figure~\ref{fig:reducer_example}(c)). The reducer in Figure~\ref{fig:reducer_example}(c) ``over-approximates''  all possible outputs of the standard deviation computation; it computes an output even if $a$ is not the average of inputs. However, the reducer is still commutative and its commutativity verification can still be done.

@@

\textbf{The Preprocessor component} first compiles a reducer program to bytecode and use the tool Soot~\cite{soot} to further convert it to the so called ``Jimple'' format, which is an intermediate language that is designed to simplify the analysis of Java programs. Under the Hadoop MapReduce framework, the permutation of the input is handled by the scheduler/shuffler component and is affected by things like network latency, which is not controllable by programmers. In order to deal with this issue, we wrote our own dummy Hadoop environment for the reducer as a part of the Preprocessor component. So that the input order of the reducer is now controlled by J-ReCoVer. Finally, we perform in the Preprocessor some program transformations tasks to simplify the analysis. For example, we convert all output commands $out(e)$ inside loops to an special assignment (Section~\ref{sec:program_trans1} and~\ref{sec:program_trans2}).

@@

\textbf{The Prover component} reduces the commutativity problem to a SMT formula satisfiability problem. At a high level, we check equivalence between a reducer program and its variant that has two consecutive inputs swapped. We show that this can be reduced to a SMT problem and give it to the SMT solver Z3~\cite{z3} for solving. In case that Z3 proved the formula unsatisfiable, we know that swapping any two consecutive inputs of the reducer will not change its output. Because all permutations of a list can be obtained by swapping consecutive elements finitely many times, it follows that the reducer outputs the same value for all permutations of the same list of inputs.
In this case,  J-ReCoVer stops and reports that the reducer is commutative. 
If we run the approach in a naive manner, it does not scale to large programs and is very imprecise.  We will introduce two critical optimizations that enables the proposed approach working for real-world examples in Section~\ref{section:optimizations}: a data flow analysis to extract critical variables and an improved symbolic execution algorithm to generate SMT formulae. More details of the algorithms used in the Preprocessor and the Prover components will be explained in Section~\ref{sec:preprocessor_prover} and the optimizations will be introduced in~\ref{section:optimizations}.

@@

Before entering the Prover component, the Preprocessor first performs a few program transformation tasks to simplify the verification task.
We assume the reducer $s_1;\rloop\{s_2\};s_3$ is given. The section begins with an introduction to the first program transformation task (Section~\ref{sec:program_trans1}) that handles the case when $s_2$ contains $out$ commands. The output of the first transformation task is a commutativity-equivalent reducer program where $out$ does not occur in $s_2$.  Then the algorithm of the Prover is introduced in Section~\ref{sec:prover}. This algorithm works only in a special case where no read from input occurs in $s_1$. For general reducer programs, we introduce the second program transformation task (Section~\ref{sec:program_trans2}) whose output is a reducer program that falls in the special case required by the Prover and ``over-approximates'' all input/output behaviors of the original reducer. The commutative of the produced reducer implies the commutative of the original one, but not vice verse.

@@

\subsection{The Prover: Reducing Commutativity to SMT Solving}
\label{sec:prover}
Below we begin with a simple version where $s_1$ never calls the $\cur$ function. Later we will explain how to handle the general case using program transformation. The command $s_2$ can be viewed as a function $F$ that reads the values of all variables and the current input before executing $s_2$ and outputs values of all variables after $s_2$. Formally, the function $F(n,x_1,x_2,\ldots,x_k): \Z^{k+1} \rightarrow \Z^k$ returns a tuple of values $x_1',x_2',\ldots,x_k'$, where $n$ is the current input value of the reducer, $x_i$ and $x_i'$ are the values of variables before and after the execution of $s_2$, respectively. The construction of $F$ from $s_2$ can be done in the standard way. We will introduce a more efficient data structure for the construction in Section~\ref{section:optimizations}.
We check if the formula below is valid for all possible values of $n_1,n_2, x_1,x_2,\ldots,x_k$. 
\begin{equation}
 F(n_1, F(n_2,x_1,x_2,\ldots,x_k)) = F(n_2, F(n_1,x_1,x_2,\ldots,x_k) )
\label{eq:commu}
\end{equation}

@@

The validity of Equation~(\ref{eq:commu}) implies that swapping any two consecutive input values does not change the final variable valuation. Informally, it says that starting from the same initial valuations of variables and with two different input orders, $[n_1;n_2]$ and $[n_2;n_1]$, the values of all program variables are the same after we execute the loop body twice. The first execution reads $n_1$ and then reads $n_2$. The other execution reads the two inputs in the reverse order.

@@

We explain why that the validity of Equation~(\ref{eq:commu}) is sufficient to prove the commutativity of the reducer $s_1;\rloop\{s_2\};s_3$. Notice that $s_1$ does not read from the input (assumed at the beginning of this subsection). The execution of $s_1$ is deterministic in the sense that from the same initial valuation, it outputs the same sequence of values and ends at the same valuation. For the execution of $\rloop\{s_2\}$, from the validity of Equation~(\ref{eq:commu}), any permutation of input values leads the reducer to the same final valuation and the same sequence of outputs at the end of the loop. 
Before executing the commands $s_3$, all inputs have been read (the loop exit condition) already. Similar the case of $s_1$, $s_3$ does not read from the input and hence the execution of $s_3$ is deterministic. It follows that for all permutations of inputs, the reducer outputs the same sequence of values.

@@

We handle the problem again by program transformation. To handle the example in Figure~\ref{fig:reducer_max} (a), we move the prefix $s_1$ into the loop body and use a new variable $s$ to force that $s_1$ is always executed before the original loop body. The result after the transformation is demonstrated in Figure~\ref{fig:reducer_max} (b). The new reducer program over-approximates all possible inputs/outputs of the original one. For the case that $s=1$ at the beginning, its output sequence is the same as the original reducer. When $s\neq 1$, it may produce outputs that are not possible by the original reducer.
It also satisfies all assumptions required for the Prover component: 
(I) the commands before entering the loop does not have any $\cur$ expression, 
(II) each branch (recall that ``branch'' is formally defined in Section~\ref{section:integer-reducers}) of the loop body has at most one $\cur$, and 
(III) $out(v)$ command does not occur inside the loop body\footnote{We need to apply the first program transformation again to fulfill this assumption.}. So we can apply Equation~(\ref{eq:commu}) to prove its commutativity.
Since the new reducer has all the possible outputs of the original one, the new reducer being commutative implies that the original one is also commutative.

\begin{figure}[htb]
	\centering
	\begin{minipage}{0.8\textwidth}

	\begin{algorithm}[H]
		$s'_0;s:=*;$\;
		\Loop{}{
			\lIf{$s=1$}{ $x_{i_1} := \cur; s'_1; s:=2$}
			\lElseIf{ $s =2$}{ $x_{i_2} := \cur; s'_2; s:=3$}
			\lElseIf{ $s = 3$}{$\ldots$}
			\lElseIf{ $s = m$}{$x_{i_m} := \cur; s'_m; s:=m+1$}
			\lElse{ $s_2$}
		};$s_3$\;
	\end{algorithm}
	\end{minipage}
	\caption{The second transformation task}
	\label{fig:general_program_transformation}
\end{figure}

For the general case, we first rewrite the $s_1$ part of the reducer $s_1;\rloop\{s_2\};s_3$ to the form 
$$s'_0;x_{i_1}{:=} \cur;s'_1.\ldots x_{i_{m}}{:=}\cur;s'_m, $$ where $\cur$ never occur in $s'_0,s'_1,\ldots, s'_m$. Then we produce the reducer in Figure~\ref{fig:general_program_transformation}, which is an over-approximation of the original reducer for the same reason: the case that $s=1$ has the same output sequence to the original reducer and when $s\neq 1$, it may produce additional behavior. Observe that the reducer in Figure~\ref{fig:general_program_transformation} satisfies all assumptions required for the Prover. Hence the Equation~(\ref{eq:commu}) can then be applied to prove its commutativity.

\section{Optimizations}
\label{section:optimizations}
In this section we explain how data flow analysis is used in J-Recover to improve the precision of commutativity checking and the algorithm we used to construct the $F$ function in Equation~(\ref{eq:commu}).

\subsection{Improving the precision of commutativity verification using data flow analysis}
\label{section:opt1}
We find that in many cases, Equation~(\ref{eq:commu}) is a condition that is too strong for the reducer to be commutative. For instance, consider the example in Figure~\ref{fig:reducer_opt}. In the loop body, the input is first stored in a variable $t$ and then added to $s$. In this case, the $F$ function returns the updated values of both $s$ and $t$ after the execution of loop body. Observe that 
$F(c_1, F(c_2,s,t)) = (c_1+c_2, c_1)$ and $F(c_2, F(c_1,s,t)) = (c_1+c_2, c_2)$. It follows that the corresponding Equation~(\ref{eq:commu}) is invalid. The second component of the returned tuples corresponds to the value of $t$ after executing the loop body twice. Their values are $c_1$ and $c_2$, respectively, in $F(c_1, F(c_2,s,t))$ and $F(c_2, F(c_1,s,t))$. However, in this case, the value of $t$ will not affect the output of this reducer.

@@

To handle such a situation, we need an algorithm to collect a set of variables whose value may affect the output. 
Recall that the first program transformation task uses a set of variables $V_2=\{v_{out_k} \mid k \in [1,l]\}$, where $l$ is the number of output commands occurred in $s_2$, to collect output values occurred in the loop body $s_2$.
The set $V_2$ is exactly the set of variables that will affect the output inside the loop.

We perform a simple backward data flow analysis to collect all variables whose value may propagate to the output command after the loop execution. With out loss of generality, we assume there is only one output command after the loop. Then the code after the loop (i.e., the $s_3$ part of the reducer) can only be in the form of Figure~\ref{fig:nested_out}.
We unfold all concatenation of commands, so now the commands $c_1\ldots c_n$ are either assignment commands $w:=e$ or branch command $\ite{g}{c'_1}{c'_2}$, and the only output command stays in the $m$-th layer of the branch command after $c_i$. For the case that $m=0$, the entire branch command in Lines 2-6 of Figure~\ref{fig:nested_out} is replaced with $out(v)$.

The data flow analysis algorithm works as follows.
A set $V_3$ whose value is initially a singleton set $\{v\}$ is used to remember important variables that might flow to the output command $out(v)$ in $s_3$ (Figure~\ref{fig:nested_out}).
The commands and guards will be inspected in the following order $c_{n}$,$\ldots$,$c_{i+1}$, $g_m$, $g_{m-1}$, $\ldots$, $g_1$, $c_i$, $\ldots$, $c_2$, $c_1$. For the case that a guard $g_j$ for $j\in [1,m]$ is being processed, we simply add all variables occurred in $g_j$ to $V_3$. 
For the case that a command $c_j$ for $j\in [1,n]$ is being processed, 
recall that $c_j$ can only be in the form of an assignment $w:=e$ or a branch $\ite{g}{c'_1}{c'_2}$.
For the case that $c_j$ is an assignment $w:=e$ and $w\in V_3$, we add all variables contained in $e$ to $V_3$. For the case that $c_j$ is a branch command $\ite{g}{c'_1}{c'_2}$, we process both $c'_1$ and $c'_2$ in a similar manner. Starting from the same set $V_3$,  after processing $c'_1$ and $c'_2$, we get two sets $V_3^1$ and $V_3^2$. We take their union $V_3^1\cup V_3^2$ as the new value of $V_3$. After the entire $s_3$ is being processed we get the set of important variables that might affect the outputs.

Then we modify the Equation~(\ref{eq:commu}) to the one below
\begin{equation}
F(n_1, F(n_2,x_1,x_2,\ldots,x_k) )=_{PosF(V_2\cup V_3)} F(n_2, F(n_1,x_1,x_2,\ldots,x_k) )
\label{eq:commu2}
\end{equation}
The function $PosF(X)$ returns the corresponding positions of variables in $X$ in the output of $F(n,x_1,\ldots,x_k) $. E.g., if $X = \{x_1,x_4,y\}$, where $y\notin \{x_1,\ldots,x_k\}$, then $PosF(X)=\{1,4\}$.
The use of equation~(\ref{eq:commu2}) is critical to the success of our approach. In our evaluation, the ratio of examples that our approach can successfully analyze is significantly increased from $6.8\%$ to $97.5\%$. The details of the evaluation can be found in Section~\ref{section:exp}.

@@

\subsection{Comparison with competing tools}
As already pointed out in~\cite{ChenHSW15}, a reducer commutativity problem can be reduced to a reachability problem of a program. 
The examples in~\cite{ChenHSW15}, which includes \textsf{rangesum}\footnote{The sum of the second half of the input list.}, \textsf{avg}, \textsf{max}, \textsf{sep}\footnote{The number of occurrences of even numbers minus that of odd numbers.}, \textsf{sum}, are already included in SV-Comp~\cite{svcomp} (Software Verification Competition) benchmarks.
Although those examples are tiny and simple, none of the software verifiers can verify all of those examples according to the 2018 results. In contrast, J-Recover can verify all them very efficiently.
